HO unification from object language to meta language

### 2024-03-30 paper/main.tex
6.1

\newcommand{\maybeeta}{\ensuremath{\Diamond\eta}\xspace}
\newcommand{\maybebeta}{\ensuremath{\overline{\llambda}}\xspace}
%\newcommand{\maybebeta}{\ensuremath{\Diamond\beta}\xspace}
\begin{definition}[\maybeeta]
  $\maybeeta = \{ t ~|~ \exists \rho, \rho t ~\mathrm{is~an~eta~expansion} \}$
\end{definition}

@@

\begin{definition}[\maybebeta]
  $\maybebeta = \{ X t_1 \ldots t_n ~|~ X t_1 \ldots t_n \not\in \llambda \}$.
\end{definition}

@@

\section{Handling of \maybeeta}\label{sec:eta}

$\eta$-reduction is an equivalence relation where a term of the form
$\lambda x.t \appsep x$ can be converted to $t$ any time $x$ does not occur as a
free variable in $t$. We call $\lambda x. t \appsep x$ the $\eta$-expansion of $t$. 
\printAll
  {{\elpiIn{flam x\ fapp [fuva X, x]} \Uo \elpiIn{fcon"f"}}}
  {{\lambda x. X'_x \Ue f}}
  {{\mapping{\elpiIn{X}}{X'}{1}}}
  {{}}
Following the implementation of \elpiIn{comp} given in
\cref{sec:compilation}, the unification problem $\mathcal{P}_1$ given 
above, is compiled into the unification problem $\mathcal{T}_1$. 
We can note that all the terms appearing in $\mathcal{P}_1$ are
unifiable by \Uo, but, despite this, \Ue, is not able to solve $\mathcal{T}_1$:
its left and right hand sides have both different rigid head:
$\lambda x.A'_x$ corresponds to the \Ho term \elpiIn{lam x\ app[con "f", x]} 
and $f$ corresponds to \elpiIn{con"f"}. 
This failure is motivated by the fact that \elpiIn{flam x\ fapp [fuva A, x]}
is a term belonging to \maybeeta.

@@

\begin{definition}[reduce-to]
  A term $t$ reduce to a name $x$, if $\exists \sigma, \sigma t = x$. In
  particular, for any term $t$, $\lambda x_1 \ldots x_n.t_{x_1\ldots x_n}$
  reduces to a bound variable $x$ if one of the three following cases is
  satisfied: 1) $n = 0$ and $t = x$; 2) $t$ is the application of $x$ to a list
  of terms $l$ and each $l_i$ reduces to $x_i$
  (e.g. $\lambda x_1 \ldots x_n.x_{x_1\ldots x_n} = x$) ; 3) $t$ is a unification
  variable in \llambda with scope
  $s$, and for any $v \in \{ x, x_1 \ldots x_n \}$,
  there exists a $s_i \in s$, such that $s_i$ reduces to $v$
\end{definition}

\begin{definition}[occurs-rigidly]
  A name $x$ occurs rigidly in a term $t$, if $\forall \sigma, x \in
  \subterm{\sigma t}$
\end{definition}
In other words $x$ \emph{occurs-rigidly} in $t$ if it occurs in $t$
outside of the scope of unification variables since theirs instantiations
are allowed to discard their scope.

@@

\begin{definition}[\maybeeta detection]
  A term $\lambda x_1 \ldots x_n.t$ where each $x_i$ occurs in $t$ is a
  \maybeeta if
  either: 1) $t$ is a constant applied to arguments
  $l_1 \ldots l_m$ such that 
  $m \geq n$ and every $l_{m-n+i}$ reduces-to $x_i$ and
  no $x_i$ occurs-rigidly in $l_{1\ldots m-n-1}$; or 2) $t$ is a
  unification variable with scope $s$ and
  for each $x_i$ there exists a $s_j \in s$ such that $s_j$ reduces-to $x_i$.
\end{definition}

As a final remark, the \maybeeta detection defined just before is an
over-approximation, in the sense that there exists some terms $t$ considered as
\maybeeta, such that forall substitution $\sigma$, $\sigma t$ is not an
$\eta$-expansion. A small example is $\lambda x. f \appsep A_x \appsep A_x$
which is considered a \maybeeta. Thi is because, we suppose at the same time
that the second $A_x$ can reduce to $x$, to the first $A_x$ is not reducible to
$A_x$. This is however not considered as a problem, since, adding holes to 
the compiled term does not break \cref{prop:simulation}, since each hole is
connected to a \linketa which is activated exactly when the hole is instantiated
the \linketa, as explain below is activated.

@@

@@

>>>>

\newcommand{\maybeeta}{\ensuremath{\Diamond\eta}\xspace}
\newcommand{\maybebeta}{\ensuremath{\overline{\llambda}}\xspace}
%\newcommand{\maybebeta}{\ensuremath{\Diamond\beta}\xspace}
\begin{definition}[\maybeeta]\label{def:maybeeta}
  $\maybeeta = \{ t ~|~ \exists \rho, \rho t ~\mathrm{is~an~eta~expansion} \}$
\end{definition}

@@

\begin{definition}[\maybebeta]\label{def:maybebeta}
  $\maybebeta = \{ X t_1 \ldots t_n ~|~ X t_1 \ldots t_n \not\in \llambda \}$.
\end{definition}

@@

\section{Handling of \maybeeta}\label{sec:eta}
\marginpar{TODO: fix syntax}
$\eta$-reduction is an equivalence relation where a term of the form
$\lambda x.t \appsep x$ can be converted to $t$ any time $x$ does not occur as a
free variable in $t$. We call $\lambda x. t \appsep x$ the $\eta$-expansion of $t$. 
\printAll
  {{\elpiIn{flam x\ fapp [fuva X, x]} \Uo \elpiIn{fcon"f"}}}
  {{\lambda x. X'_x \Ue f}}
  {{\mapping{\elpiIn{X}}{X'}{1}}}
  {{}}
Following the implementation of \elpiIn{comp} given in
\cref{sec:compilation}, the unification problem $\mathcal{P}_1$ given 
above, is compiled into the unification problem $\mathcal{T}_1$. 
We can note that all the terms appearing in $\mathcal{P}_1$ are
unifiable by \Uo, but, despite this, \Ue, is not able to solve $\mathcal{T}_1$:
its left and right hand sides have both different rigid head:
$\lambda x.A'_x$ corresponds to the \Ho term \elpiIn{lam x\ app[con "f", x]} 
and $f$ corresponds to \elpiIn{con"f"}. 
This failure is motivated by the fact that \elpiIn{flam x\ fapp [fuva A, x]}
is a term belonging to \maybeeta.

@@

\marginpar{the compiler must do beta normal!!!!}
\newcommand{\reduceto}{\emph{may-contract-to}}
\begin{definition}[\reduceto]
  A term $s$ \reduceto{} a name $x$ if there exists a
  substitution $\rho$ such that $\rho s \Eo{} x$.
\end{definition}

\begin{lemma}\label{lem:reduceto}
A $\beta$-normal term $s = \lambda x_1 \ldots x_n.t$
%, where $x, x_1\ldots x_n$ can occur in $t$, 
\reduceto{} $x$ only if one of the following three conditions holds:
\begin{enumerate}
  \item $n = 0$ and $t = x$;
  \item $t$ is the application of $x$ to a list
     of terms $l$ and each $l_i$ \reduceto{} $x_i$
     (e.g. $\lambda x_1 \ldots x_n.x \appsep x_1 \ldots x_n \Eo{} x$) ;
  \item $t$ is a unification variable with scope
    $W$, and for any $v \in \{ x, x_1 \ldots x_n \}$,
    there exists a $w_i \in W$, such that $w_i$ \reduceto{} $v$
    (if $n = 0$ this is equivalent to $x \in W$).
\end{enumerate}
\end{lemma}
\begin{proof}[Proof sketch]
Since our terms are in $\beta$-normal form there is only
one rule that can play a role (namely \ref{clause:eta1}),
hence if the term $s$ is not exactly $x$ (case 1)
it can only be an $\eta$-expansion of it, or a unification
variable that can be assigned to it, or a combination of both.
If $s$ begins with a lambda, then the lambda can only disappear by $\eta$
contraction. In that case the term $t$ under the spine of binders for
$x_1\ldots x_n$ can either be $x$ itself applied to terms that can
\reduceto{} these variables (case 2), or a unification variable
that can be assigned to that application (case 3).
\end{proof}

\noindent
Note that this condition does not require the term to be in \llambda.
\marginpar{Is this relevant}

\newcommand{\occursrigid}{\emph{occurs-rigidly}}
\begin{definition}[\occursrigid]\label{def:occrigid}
  A name $x$ \occursrigid{} in a $\beta$-normal term $t$, if $\forall \rho, x \in
  \subterm{\rho t}$
\end{definition}
In other words $x$ \emph{\occursrigid} in $t$ if it occurs in $t$
outside of the scope of unification variables since an instantiation
is allowed to discard $x$ from the scope of the unification variable.
Note that $\eta$-contraction cannot make $x$ disappear, since the
variables being erased by $\eta$-contraction are locally bound inside $t$.

@@

\begin{lemma}[\maybeeta detection]\label{lem:maybeeta}
  A $\beta$-normal term
  $\lambda x_1 \ldots x_n.t \in \maybeeta$ (as per \cref{def:maybeeta})
  % where each $x_i$ occurs in $t$ is a
  %\maybeeta if
  %either: 
  if one of the following holds:
  \begin{enumerate}
    \item $t$ is a constant or variable applied to arguments
      $l_1 \ldots l_m$ such that 
      $m \geq n$ and for every $i$ such that $1 \leq i \leq m-n$
      the term  $l_i$
      \reduceto{} $x_i$, and
      no $x_i$ \occursrigid{} in $l_1 \ldots l_{m-n-1}$;
    \item $t$ is a
      unification variable with scope $W$ and
      for each $x_i$ there exists a $w_j \in W$ such that $w_j$
      \reduceto{} $x_i$.
  \end{enumerate}
\end{lemma}
\begin{proof}[Proof sketch]
Follows from \cref{def:occrigid} and \cref{lem:reduceto}
\end{proof}

\noindent
Remark that the converse of \cref{lem:maybeeta} does not hold: 
there exists a term $t \in \maybeeta$
such that forall substitution $\rho$, $\rho t$ is not an
$\eta$-expansion. A simple counter example is
$\lambda x. f \appsep (A \appsep x) \appsep (A \appsep x)$
which is in \maybeeta{} since $x$ does not \occursrigid{} in the first argument
of $f$,
and the second argument of $f$ \reduceto{} $x$. In other words $A \appsep x$
may either use $x$ or discard it.\\
This is however not a problem in the sense that it does not break
\cref{prop:simulation}. Terms in \maybeeta{} are compiled to
unification variables and a link (see \cref{sec:etacomp}):
the link makes progress (see \cref{sec:etaprogress})
in the same step in which the
variable is instantiated, and that compensates for the
potential misclassification.

@@

\subsection{Compilation}\label{sec:etacomp}

@@

\subsection{Progress}\label{sec:etaprogress}
